---
title: 'Some figures for chapter 1'
author: "Haakon Bakka"
date: "2023 August 11"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    number_sections: yes
---


## Description

This Rmarkdown produces some of the plots in Chapter 1 of Hrafnkelsson et al.


```{r}
stopifnot(file.exists("Book-ch1-figs.Rmd"))
folder.out = "fig-out/"
dir.create(folder.out, showWarnings = F)
```

```{r}
library(boot)
library(ggplot2)
library(Matrix)
library(INLA)
```



## Figure 1

The flat likelihoods:

```{r}
library(boot)
etavals = seq(-12, 2, length.out = 300) 
gtrue = dbinom(x = 0, size=4, prob=inv.logit(etavals), log = T) 
plot(etavals, gtrue, xlab="eta", ylab="Log-likelihood")

png(paste0(folder.out, "likelihood-flat.png"))
etavals = seq(-12, 2, length.out = 300) 
gtrue = dbinom(x = 0, size=4, prob=inv.logit(etavals), log = T) 
plot(etavals, gtrue, xlab="Linear predictor", ylab="Log-likelihood")
dev.off()
```

```{r}
## GGPLOT
df = data.frame(etavals, gtrue)
# Basic line plot with points
ggp.lik = ggplot(data=df, aes(x=etavals, y=gtrue))  + theme_bw() +
          geom_line(size=2) + xlab("Linear predictor") + ylab("Log-likelihood")
ggp.lik
ggsave(ggp.lik, filename = paste0(folder.out, "likelihood-flat.png"), height=5, width=7)
```



## Figure 2


We add a normal(0,1) prior, and do a quadratic approximation at the mode.

```{r}
N.trials = 4
prior.sd = 1
for (i in 1:3) {
  if (i==1) {
    yval = 2
    eta.star = 0 # is the logit(yval/Ntrials) and is the mean of prior
  } else if (i==2) {
    yval = 1
    eta.star = -0.5
  } else {
    yval = 0
    eta.star = -1.05
  }
  p = inv.logit(eta.star)
  lik.d0 = dbinom(x=yval, prob = p, size=N.trials, log=T)
  lik.d1 = yval-N.trials*p
  lik.d2 = - N.trials*p*(1-p)
  a = lik.d0 - eta.star*lik.d1 + 0.5*eta.star^2*lik.d2
  b = lik.d1 - eta.star*lik.d2
  c = -lik.d2
  etavals = seq(-1, 1, length.out = 300) + eta.star
  gapprox = a + b*etavals - 1/2 * c *etavals^2
  gtrue = dbinom(x = yval, size=N.trials, prob=inv.logit(etavals), log = T) 
  ## Multiply both with the prior
  gapprox = gapprox + dnorm(x=etavals, log = T)
  gtrue = gtrue + dnorm(x=etavals, log = T)
  # = p.and.d12.y.given.eta(eta = eta.star, order = 0)
  plot(etavals, gtrue, type="l", lwd=2, ylab="Log probability", xlab="Linear predictor")
  lines(etavals, gapprox, col="blue", lwd=2, lty="dashed")
  abline(v=eta.star)
  
  png(paste0(folder.out, "laplace-withg-y",yval, ".png"))
  plot(etavals, gtrue, type="l", lwd=2, ylab="Log probability", xlab="Linear predictor")
  lines(etavals, gapprox, col="blue", lwd=2, lty="dashed")
  abline(v=eta.star)
  dev.off()
  
}
```

Pick one of the plots and use ggplot

```{r}
## GGplot
df = data.frame(etavals, gtrue, gapprox)
# Basic line plot with points
ggp.lik = ggplot(data=df, aes(x=etavals, y=gtrue))  + theme_bw() +
  geom_line(size=2) +
  #geom_line(aes(y=gapprox), size=2, lty="dotted", col="blue") +
  geom_line(aes(y=gapprox), size=1, col="blue") +
  xlab("Linear predictor") + ylab("Log probability")
ggsave(ggp.lik, filename=paste0(folder.out, "laplace-withg-y",yval, ".png"), height=5, width=7)
ggp.lik
```





## Figure 11


```{r}
## Note to Birgir: This data is created by the other .rmd file
load("precip.Rdata")
```

```{r}
head(precip)
```
The first column is the day in the year, and the second is a count of how many of the four years where there was precipitation on that day.
We deleted the 29th of February year 2020.

We can plot the data as follows.

```{r}
plot(precip$day, precip$precip, ylab="Days with precipitation", xlab="Day in calendar year")
png(paste0(folder.out, "precip.png"))
plot(precip$day, precip$precip, ylab="Days with precipitation", xlab="Day in calendar year")
dev.off()
```

```{r}
## GGPLOT
ggp.precip = ggplot(data=precip, aes(x=day, y=precip))  + theme_bw() +
          geom_point(size=1) + xlab("Day in calendar year") + ylab("Days with precipitation")
ggsave(ggp.precip, filename = paste0(folder.out, "precip.png"), height=5, width=7)
ggp.precip
```




## Figure 12

```{r}
ma <- function(x, n = 5){filter(x, rep(1 / n, n), sides = 2)}
m1 = ma(precip$precip, n=5)
m2 = ma(precip$precip, n=20)
plot(m1, ylab="Moving average precipitation", xlab="Day in calendar year", lwd=2)
lines(m2, col="blue", lwd=2)
png(paste0(folder.out, "precip-ma.png"))
plot(m1, ylab="Moving average precipitation", xlab="Day in calendar year", lwd=2)
lines(m2, col="blue", lwd=2)
dev.off()
```
```{r}
## GGPLOT
df2 = data.frame(day=1:365, m1, m2)
ggp.precip = ggplot(data=df2, aes(x=day, y=m1))  + theme_bw() +
          geom_line(size=1) + 
          geom_line(aes(y=m2), col="blue") + 
  xlab("Day in calendar year") + ylab("Moving average precipitation")
ggsave(ggp.precip, filename = paste0(folder.out, "precip-ma.png"), height=5, width=7)
ggp.precip
```



```{r}
barplot(as.numeric(m1))
barplot(as.numeric(m2))
```




## Figure 13
A circular auto-regressive model component.

```{r}
N = nrow(df2)
#N = 10
## For example with
rho = 0.95
Ui = sparseMatrix(1:N, 1:N, x = rep(1, N), dims = c(N,N))
Uim1 = sparseMatrix(2:N, 1:(N-1), x = rep(1, N-1), dims = c(N,N))
T1 = Ui - rho*Uim1
T2 = T1 %*% T1
Q2 = t(T2) %*% T2

```

```{r}
## Make circular
Q2[1, N] = Q2[1, 2]
Q2[1, N-1] = Q2[1, 3]
Q2[2, N] = Q2[2, 4]
Q2[N, 1] = Q2[1, N]
Q2[N-1, 1] = Q2[1, N-1]
Q2[N, 2] = Q2[2, N]

#Q2[1:2, 1:2] = Q2[N - 0:1, N - 0:1]
Q2[N - 0:1, N - 0:1] = Q2[1:2, 1:2]

if(N<11) print(round(Q2, 3))

```

```{r}
## Fix variance problem
range(diag(solve(Q2)))
phi1 = 2*rho
phi2 = -rho^2
mvar = (1-phi2)/((1+phi2)*(1-phi1-phi2)*(1+phi1-phi2))
Q3 = Q2 * mvar
range(diag(solve(Q3)))
const.corr = mean(diag(solve(Q2)))
Q3 = Q2*const.corr
range(diag(solve(Q3)))
```



```{r}
## Correlation
Sig3 = solve(Q3)
Sig3[1, 91]
Sig3[90, 90]
```



```{r}
T2 = chol(Q3)
u2.sim = solve(T2, rnorm(N))
plot(u2.sim)
```

```{r}
set.seed(202106)

dftemp = as.data.frame(as.matrix(solve(T2, cbind(rnorm(N), rnorm(N), rnorm(N), rnorm(N), rnorm(N), rnorm(N)))))
dftemp$day = df2$day
ggp.sim = ggplot(data=dftemp, aes(x=day, y=V1)) + theme_bw() +
          geom_line(aes(y=V5), col="black") + 
          geom_line(aes(y=V2), col="blue") + 
          geom_line(aes(y=V3), col="red") + 
          geom_line(aes(y=V4), col="darkgreen") + 
          xlab("Day in calendar year") + ylab("Simulated u")
ggp.sim

ggsave(ggp.sim, filename = paste0(folder.out, "sim-u.png"), height=5, width=7)
```




```{r}
Qar2c = Q3
```



## Figure 14: Inference

We will use the R-package INLA to perform inference with the Laplace approximation.

```{r}
library(INLA)
hyper1 = list(theta = list(prior="pc.prec", param=c(0.1,0.5)))
## Choose formula
if (T) {
form1 = precip ~ 1 + f(day, model="generic0", Cmatrix = Qar2c, hyper=hyper1)
} else {
  ## Old version
form1 = precip ~ 1 + f(day, model="rw2", constr=TRUE, cyclic=TRUE, scale.model=TRUE, hyper=hyper1)
}


```

We removed the intercept since this is undetermined in the RW2 model (it is in the null space).
We also used the cyclic effect $u_1 = u_{365}$

```{r}
fit1 = inla(form1, data=precip, family="binomial", Ntrials = 4,
            control.predictor=list(compute=T),
            control.inla = list(int.strategy="eb", strategy="gaussian"))
```

```{r}
summary(fit1)
```

```{r}
plot(fit1$summary.random$day$ID, fit1$summary.random$day$mean,
     xlab="Day", ylab="Logit probability contribution")
```

```{r}
plot(1:365, fit1$summary.fitted.values$mean,
     xlab="Day", ylab="Probability of precipitation")
```
### Intermezzo: Individual beta prior

```{r}
## Assume beta(1,1) prior, we get Beta:
post.b1 = 1+precip$precip
post.b2 = 1+4-precip$precip
post.m = post.b1/(post.b1+post.b2)
q1 = qbeta(0.5, post.b1, post.b2)
q2 = qbeta(0.025, post.b1, post.b2)
q3 = qbeta(0.975, post.b1, post.b2)
q = rbind(q1, q2, q3)
```




Plot together with data: Measured and expected number of events.
```{r}
plot(precip$day, precip$precip, xlab="Day", ylab="Precipitation")
lines(1:365, 4*fit1$summary.fitted.values$mean, col="blue")
lines(1:365, q1*4, col="red")
lines(1:365, q2*4, col="red", lty="dotted")
lines(1:365, q3*4, col="red", lty="dotted")

```

This plot is the goal of this analysis. 
It shows increased precipitation around day 30, decreased precipitation around day 90, and increased precipitation later in the year.

### More advanced approximations
There are two main steps to improving the above approximation.
The first is to integrate over the $\theta$ posterior.
This can be done either with sampling (in MCMC algorithms) or with deterministic integration (in INLA).
The second is to improve the estimates of the posterior marginals of $u$, namely $\pi(u_i|y)$, by another Laplace approximation, see INLA review paper.



```{r}
fit2 = inla(form1, data=precip, family="binomial", Ntrials = 4,
            control.predictor=list(compute=T))
```

This code is slightly simpler than for `fit1`, because this is the default option in INLA.

Next we compare the old and new results, and the quantiles estimated from the marginals.
```{r}
mult = 1
#ylab1 = "Expected days with precipitation"
ylab1 = "Probability of precipitation"
plot(1:365, mult*fit1$summary.fitted.values$mean, type="l", ylim=c(0.1, 0.9)*mult,
     ylab=ylab1, xlab="Day in calendar year")
lines(1:365, mult*fit1$summary.fitted.values$"0.025quant", type="l")
lines(1:365, mult*fit1$summary.fitted.values$"0.975quant", type="l")
lines(1:365, mult*fit2$summary.fitted.values$mean, col="blue", cex=0.5)
lines(1:365, mult*fit2$summary.fitted.values$"0.025quant", col="blue", cex=0.5)
lines(1:365, mult*fit2$summary.fitted.values$"0.975quant", col="blue", cex=0.5)
#lines(1:365, q1*mult, col="red", type="p")
#lines(1:365, q2*mult, col="red", lty="dotted")
#lines(1:365, q3*mult, col="red", lty="dotted")
#points(precip$day, precip$precip, col="darkgreen")

png(paste0(folder.out, "precip-est2.png"), width = 240*2, height=240*2)
plot(1:365, mult*fit1$summary.fitted.values$mean, type="l", ylim=c(0.2, 0.68)*mult,
     ylab=ylab1, xlab="Day in calendar year")
lines(1:365, mult*fit1$summary.fitted.values$"0.025quant", type="l")
lines(1:365, mult*fit1$summary.fitted.values$"0.975quant", type="l")
lines(1:365, mult*fit2$summary.fitted.values$mean, col="blue", cex=0.5)
lines(1:365, mult*fit2$summary.fitted.values$"0.025quant", col="blue", cex=0.5)
lines(1:365, mult*fit2$summary.fitted.values$"0.975quant", col="blue", cex=0.5)
dev.off()

```
```{r}
## GGPLOT
df3 = data.frame(day=1:365, e11 = mult*fit1$summary.fitted.values$mean,
                 e12 = fit1$summary.fitted.values$"0.025quant",
                 e13 = fit1$summary.fitted.values$"0.975quant",
                 e21 = fit2$summary.fitted.values$mean,
                 e22 = fit2$summary.fitted.values$"0.025quant",
                 e23 = fit2$summary.fitted.values$"0.975quant")
ggp.precip = ggplot(data=df3, aes(x=day, y=e11)) + theme_bw() +
          geom_line(aes(y=e11)) + 
          geom_line(aes(y=e12)) + 
          geom_line(aes(y=e13)) + 
          geom_line(aes(y=e21), col="blue") + 
          geom_line(aes(y=e22), col="blue") + 
          geom_line(aes(y=e23), col="blue") + 
          xlab("Day in calendar year") + ylab("Probability of precipitation")
    
ggsave(ggp.precip, filename = paste0(folder.out, "precip-est.png"), height=5, width=7)
ggp.precip
```




We note that the mean estimate is almost exactly the same, but that the upper and lower quantiles are slightly different.
This is because the posterior marginals are slightly skewed, which is not picked up on by the simple quadratic ("Gaussian") approximation.

```{r}
## Day 100, the uncertainties and teh datapoint
## DOnt plot just show all intervals
day = 101
print(precip$precip[day]/4)
f1s = c(fit1$summary.fitted.values$mean[day],
  fit1$summary.fitted.values$"0.025quant"[day],
  fit1$summary.fitted.values$"0.975quant"[day])
f2s = c(fit2$summary.fitted.values$mean[day],
  fit2$summary.fitted.values$"0.025quant"[day],
  fit2$summary.fitted.values$"0.975quant"[day])
a = data.frame(observ = c(precip$precip[day]/4, NA, NA),
               single=q[, day], laplace = f1s, inla = f2s)
rownames(a) = c("Median", "Lower", "Upper")
round(a[c(1,3,2), ], 3)
```

### Check that the model is cyclic

```{r}
ids = c(100:N, 1:100)
plot(fit2$summary.fitted.values$mean[ids], type="l", xlab="reindexed")
abline(v=N-100.5, col="blue")
```




## End


